Hippocampal segmentation with MAGeT

### 2012-11-30 paper.Rnw
Update with full ADNI baseline info; Outlined rational for experiments

\SweaveOpts{echo = FALSE, message = FALSE, warning = FALSE, fig.align = 'center', tidy = FALSE, comment = NA, cache = TRUE}
<<loadpackages, results="hide", echo=FALSE, cache=FALSE>>=
library(ggplot2)
library(Hmisc)
library(reshape2)
library(ADNIMERGE)
options(digits=3)
@

@@

\begin{abstract}
Neuroimaging research often relies on automated anatomical segmentations of MR
images of the brain. Current multi-atlas based approaches provide accurate
segmentations of brain images by propagating manually derived segmentations of
specific neuroanatomical structures to unlabelled data. These approaches often
rely on a large number of such manually segmented atlases that take significant
time and expertise to produce. We present an algorithm for the automatic
segmentation of the hippocampus that minimizes the number of atlases needed
while still achieving similar accuracy to multi-atlas approaches.
\mc{Finish this... }
\end{abstract}

@@

@@

<<ADNI-baseline-volumes-prep,echo=F,cache=T>>=
data = read.csv('data/ADNI_baseline_volumes.csv')
unscaled=subset(data,Scaled==0 & Baseline==1 & DX != '')
attach(unscaled)
totals = data.frame(LONIUID,L+R,UCSFFSL_L+UCSFSL_R,MAPER_L+MAPER_R,UCSFSNT_L + UCSFSNT_R,SNT_L + SNT_R,DX)
colnames(totals) <- c("ID","MAGeT","FSL","MAPER","UCSFSNT","SNT","DX")
@

\begin{figure}[h]
<<ADNI-baseline-volumes-boxplot,echo=F,fig=T, fig.align='center',dependson='ADNI-baseline-volumes-prep',cache=T>>=
qplot(DX,value,data=melt(totals,id.vars=c("ID","DX"),measure.vars=c("FSL","MAPER","MAGeT","SNT"),na.rm=T),
      colour=variable,geom="boxplot") + 
        xlab("Diagnosis") + ylab("Total Hippocampal Volume (mm3)")

@@

\begin{figure}[h]
<<ADNI-baseline-volumes-plot,echo=F,fig=T, fig.align='center',dependson='ADNI-baseline-volumes',cache=T>>=
by_snt_volumes = melt(totals,id.vars=c("ID","SNT"), measure.vars=c("FSL","MAPER","MAGeT"))
fsl=subset(by_snt_volumes, variable=="FSL")
maper=subset(by_snt_volumes, variable=="MAPER")
maget=subset(by_snt_volumes, variable=="MAGeT")

@@

\begin{figure}[h]
<<SZ_volumes,echo=F,fig=T, fig.align='center',dependson='ADNI-baseline-volumes',cache=T>>=
SZ_volumes=read.csv("data/SZ_volumes_by_method.csv")
qplot(Manual,MAGeT,data=SZ_volumes) + geom_smooth(method="lm") + 
#   geom_text(aes(
#       label=paste("R=", round(cor(Manual,MAGeT, use="complete.obs"), 2), 
#                   ", p<", t.test(Manual, MAGeT, paired=T)["p.value"]),
#       x=6500,y=10000, hjust=0,size=2)) +
    xlab("Total Hippocampal Volume (Manual Segmentation)") + 
    ylab("Total Hippocampal Volume (MAGeT Segmentation)")

@
  \caption{{\bf First Episode Schizophrenic Patients.} Comparison of total HC volumes for MAGeT against manually rated volumes of}
  \label{SZ_volumes}
\end{figure}
\end{document}

>>>>

\SweaveOpts{echo = FALSE, message = FALSE, warning = FALSE, fig.align = 'center', tidy = FALSE, comment = NA, cache = TRUE}
<<loadpackages, results="hide", echo=FALSE, cache=FALSE>>=
library(ggplot2)
library(Hmisc)
library(reshape2)
library(ADNIMERGE)
library(plyr)
options(digits=3)
theme_set(theme_bw())
@

@@

\begin{abstract}
Neuroimaging research often relies on automated anatomical segmentations of MR
images of the brain. Current multi-atlas based approaches provide accurate
segmentations of brain images by propagating manually derived segmentations of
specific neuroanatomical structures to unlabelled data. These approaches often
rely on a large number of such manually segmented atlases that take significant
time and expertise to produce. We present an algorithm for the automatic
segmentation of the hippocampus that minimizes the number of atlases needed
while still achieving similar accuracy to multi-atlas approaches.
\todo{Finish this... }
\end{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%                         Experiment Rationale 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\subsubsection{Study Method Rationale - An Outline}
\begin{description}
\item[ADNI Repeated Random Subsampling Validation] \hfill \\ 
{\bf Purpose:} to show method leads to improved accuracy (over multi-atlas method), and is robust
w.r.t. registration method, atlas and template library composition and size, and various fusion methods. Additionally, to choose the best configuration of parameters (under the assumption that performance generalises). 
\\ \\ 
{\bf Method:} using a set of 69 ADNI subjects, 13 from each disease category, perform 
10 rounds of repeated random subsampling to pick atlas library and template library 
for each parameter setting and then run MAGeT brain to segment non-atlas subjects.
Also perform multi-atlas and single-atlas segmentation using the same parameter settings
and atlas library size.
\\ \\
{\bf Results:} find significant improvement over multi-atlas performed with the same
parameters. Also, find smoothed performance is monotonically increasing but asymptotic
in size of both template and atlas library, with peak performance reached after 15
templates. \todo{Can we statistically capture "peak" performance?  The point at which 
gains become statistically insignificant?}

\item[Winterburn-Atlas ADNI Baseline Segmentation] \hfill \\
{\bf Purpose:} given that MAGeT is robust and accurate, show MAGeT performance on 
a real-world task in which atlas set may not be representative of subjects.
\\ \\
{\bf Method:} We use best settings from first experiment (ANTS, majority vote, 
template library of 20), and use Winterburn hi-res atlases to segment all ADNI Baseline 
images.  Since our segmentation protocol differs from what can be segmented at 
ADNI resolution a direct comparison is not appropriate.  Instead, we validate by 
showing how well our segmentation volumes correlate with existing automated/manual 
segmentation volumes. \todo{Also show disease prediction?}
\\ \\
{\bf Results: }

\item[First Episode SZ Patients Segmentation] \hfill \\ 
{\bf Purpose:} validate that MAGeT's performance generalises to other diseases.
\\ \\
{\bf Method:} Measure performance using best parameter settings on a different disease dataset. We use FEP SZ patients and controls, using two different atlas sets: a manual hippocampal segmentation of patients, and Winterburn atlas set.  We validate the native-atlas set segmentations using Dice's Kappa, and the Winterburn-atlas-set segmentations by correlating volumes. 
\\ \\
{\bf Results:} High volume correlation between Winterburn segmentation volumes and ground truth.  (High-ish?) Kappa when using manual segmentations as Atlases.

\item[Winterburn Atlas Reliability] \hfill \\
{\bf Purpose:} validation of performance on hippocampal subregion segmentation
\\ \\
{\bf Method:} we use the Winterburn atlas set, using each high-res atlas in turn 
to segment a standard-res MR of the same subject and compare the automatic 
segmentation to the atlas segmentation.  We use a template library consisting 
of standard-res MRs from the remaining Winterburn atlas subjects plus 11 other 
standard-res healthy control subjects. 
\\ \\
{\bf Results:} TBD

\end{description}
\end{kframe}

@@

<<ADNI-baseline-volumes-prep,echo=F,cache=T>>=
yr1   = read.csv("data/ADNI_baseline_volumes/ADNI1_Complete_1Yr_1.5T_11_15_2012.csv")
mb    = read.csv("data/ADNI_baseline_volumes/ADNI1_1.5T_MAGeT_volumes.csv")
maper = read.csv("data/ADNI_baseline_volumes/MAPER_volumes.csv")
snt   = read.csv("data/ADNI_baseline_volumes/UCSFSNTVOL.csv")
ucsd  = read.csv("data/ADNI_baseline_volumes/UCSDVOL.csv")
fs    = read.csv("data/ADNI_baseline_volumes/UCSFFSX_02_15_12.csv")
fsl   = read.csv("data/ADNI_baseline_volumes/ADNI1_1.5T_FSLFIRST_volumes.csv")

# recode VISCODE

viscode_levels = list("bl"=1,"bl"=2, "m06"=3,"m12"=4)
yr1$VISCODE <- as.factor(yr1$Visit)
levels(yr1$VISCODE) <-viscode_levels

yr1$Scaled_2 <- grepl("Scaled_2", yr1$Description)

subjects = subset(yr1, VISCODE == 'bl' & Scaled_2 == FALSE)

# fetch just the columns we need, and do a little renaming
mb_pruned     = data.frame(MAGeT_L = mb$Left_Hippo, MAGeT_R = mb$Right_Hippo, Source = mb$SourceImageID)
maper_pruned  = data.frame(MAPER_L = maper$Left_Hippo, MAPER_R = maper$Right_Hippo, Source = maper$SourceImageID)
snt_pruned    = data.frame(SNT_L = snt$LEFTHIPPO, SNT_R = snt$RIGHTHIPPO, Source = snt$IMAGEUID)
ucsd_pruned   = data.frame(RID=ucsd$RID, VISCODE=ucsd$VISCODE, UCSD_L=ucsd$LHIPPOC, UCSD_R=ucsd$RHIPPOC)
fs_pruned     = subset(fs, TEMPQC != "Fail")
fs_pruned     = data.frame(FS_L=fs_pruned$ST29SV, FS_R=fs_pruned$ST88SV, Source = fs_pruned$IMAGEUID)
fsl_pruned    = data.frame(FSL_L=fsl$X17, FSL_R=fsl$X53, Source=fsl$Source)

# diagnoses
dx = adnimerge[c("DX.bl", "RID", "VISCODE")]

# Now create one data.frame with all the yr1 volume data we have from each datasource
combined = merge(subjects, mb_pruned   , by.x="Image.Data.ID", by.y="Source", all.x=TRUE)
combined = merge(combined, maper_pruned, by.x="Image.Data.ID", by.y="Source", all.x=TRUE)  # only Baseline
combined = merge(combined, snt_pruned  , by.x="Image.Data.ID", by.y="Source", all.x=TRUE)
combined = merge(combined, ucsd_pruned , by=c("RID", "VISCODE"), all.x =TRUE)              # TODO: this a unique key?
combined = merge(combined, fs_pruned   , by.x="Image.Data.ID", by.y="Source", all.x=TRUE)
combined = merge(combined, fsl_pruned  , by.x="Image.Data.ID", by.y="Source", all.x=TRUE)
combined = merge(combined, dx          , by=c("RID","VISCODE"), all.x=TRUE)

attach(combined)
vols = data.frame(RID = RID, VISCODE = VISCODE, DX = DX.bl,
                  MAGeT = MAGeT_L + MAGeT_R, 
                  MAPER = MAPER_L + MAPER_R, 
                  SNT   = SNT_L + SNT_R, 
                  UCSD  = UCSD_L + UCSD_R,
                  FS    = FS_L + FS_R, 
                  FSL   = FSL_L + FSL_R)
write.csv(vols, "data/ADNI_baseline_volumes_combined_volumes.csv")
@

\begin{figure}[h]
<<ADNI-baseline-volumes-boxplot,echo=F,fig=T, fig.align='center',dependson='ADNI-baseline-volumes-prep',cache=T>>=
totals = read.csv("data/ADNI_baseline_volumes_combined_volumes.csv")
qplot(DX,value,data=melt(totals,measure.vars=c("FS", "FSL","UCSD", "MAPER","MAGeT","SNT"),na.rm=T),
      colour=variable,geom="boxplot") + 
        xlab("Diagnosis") + ylab("Total Hippocampal Volume (mm3)")

@@

\begin{figure}[h]
<<ADNI-baseline-volumes-plot,echo=F,fig=T, fig.align='center',dependson='ADNI-baseline-volumes-boxplot',cache=T>>=
by_snt_volumes = melt(totals, measure.vars=c("FSL","MAPER","MAGeT", "FS", "UCSD"))
fsl=subset(by_snt_volumes, variable=="FSL")
maper=subset(by_snt_volumes, variable=="MAPER")
maget=subset(by_snt_volumes, variable=="MAGeT")

@@

\begin{figure}
<<SZ_volumes,echo=F,fig=T, fig.align='center',cache=T>>=
SZ_volumes=read.csv("data/SZ_volumes_by_method.csv")
qplot(Manual,MAGeT,data=SZ_volumes) + geom_smooth(method="lm") + 
#   geom_text(aes(
#       label=paste("R=", round(cor(Manual,MAGeT, use="complete.obs"), 2), 
#                   ", p<", t.test(Manual, MAGeT, paired=T)["p.value"]),
#       x=6500,y=10000, hjust=0,size=2)) +
    xlab("Total Hippocampal Volume (Manual Segmentation)") + 
    ylab("Total Hippocampal Volume (MAGeT Segmentation)")

@
  \caption{{\bf First Episode Schizophrenic Patients.} Comparison of total HC volumes for MAGeT against manually rated volumes of}
  \label{SZ_volumes}
\end{figure}

Some other text here... 
\end{document}

### 2012-11-30 paper.tex
Calculating r/p-values for ADNI/SZFEP plots, but display code currently commented out due to ugliness
Update with full ADNI baseline info; Outlined rational for experiments

@@

Age~at~baseline~\hfill\tiny{Years}&{\scriptsize 72.20~}{75.50 }{\scriptsize 78.45} &{\scriptsize 71.00~}{77.10 }{\scriptsize 81.40} &{\scriptsize 71.70~}{77.80 }{\scriptsize 81.75} &{\scriptsize 71.50~}{76.60 }{\scriptsize 81.30} &$ F_{2,66}=0.15 ,~ P=0.862 ^{1} $\NN

@@

CDR-SB&{\scriptsize 0.00~}{0.00 }{\scriptsize 0.00} &{\scriptsize 0.75~}{1.50 }{\scriptsize 1.50} &{\scriptsize 4.00~}{4.50 }{\scriptsize 5.00} &{\scriptsize 0.00~}{1.50 }{\scriptsize 4.00} &$ F_{2,66}=294.2 ,~ P<0.001 ^{1} $\NN
ADAS~13&{\scriptsize  4.665~}{ 5.670 }{\scriptsize 12.335} &{\scriptsize 14.335~}{16.000 }{\scriptsize 20.500} &{\scriptsize 23.830~}{29.000 }{\scriptsize 31.665} &{\scriptsize 10.000~}{16.000 }{\scriptsize 25.330} &$ F_{2,66}=109.8 ,~ P<0.001 ^{1} $\NN
MMSE&{\scriptsize 28.5~}{29.0 }{\scriptsize 30.0} &{\scriptsize 25.0~}{27.0 }{\scriptsize 28.0} &{\scriptsize 21.0~}{23.0 }{\scriptsize 24.0} &{\scriptsize 24.0~}{27.0 }{\scriptsize 29.0} &$ F_{2,66}=94.17 ,~ P<0.001 ^{1} $\LL

@@

@@

  \caption{{\bf First Episode Schizophrenic Patients.} Comparison of total HC volumes for MAGeT against manually rated volumes of}
  \label{SZ_volumes}
\end{figure}
\end{document}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%                         Experiment Rationale 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\subsubsection{Study Method Rationale - An Outline}
\begin{description}
\item[ADNI Repeated Random Subsampling Validation] \hfill \\ 
{\bf Purpose:} to validate robustness of MAGeT accuracy when atlas set is representative 
of subject set w.r.t. registration method, atlas and template library composition and size, and 
various fusion methods. Additionally, to choose the best configuration of parameters 
(under the assumption that performance generalises). 
\\ \\ 
{\bf Method:} using a set of 69 ADNI subjects, 13 from each disease category, perform 
10 rounds of repeated random subsampling to pick atlas library and template library 
for each parameter setting and then run MAGeT brain to segment non-atlas subjects.
Also perform multi-atlas and single-atlas segmentation using the same parameter settings
and atlas library size.
\\ \\
{\bf Results:} find significant improvement over multi-atlas performed with the same
parameters. Also, find smoothed performance is monotonically increasing but asymptotic
in size of both template and atlas library, with peak performance reached after 15
templates. \todo{Can we statistically capture "peak" performance?  The point at which 
gains become statistically insignificant?}

\item[Winterburn Atlas Reliability] \hfill \\
{\bf Purpose:} validation of reliability of whole hippocampus and hippocampal 
subregion segmentations.
\\ \\
{\bf Method:} we use the Winterburn atlas set, using each high-res atlas in turn 
to segment a standard-res MR of the same subject and compare the automatic 
segmentation to the atlas segmentation.  We use a template library consisting 
of standard-res MRs from the remaining Winterburn atlas subjects plus 11 other 
standard-res healthy control subjects. 
\\ \\
{\bf Results:} TBD

\item[Winterburn-Atlas ADNI Baseline Segmentation] \hfill \\
{\bf Purpose:} given that MAGeT is robust and accurate, show MAGeT performance on a real-world task in which atlas set may not be representative of subjects.
\\ \\
{\bf Method:} We use best settings from first experiment (ANTS, majority vote, 
template library of 20), and use Winterburn hi-res atlases to segment all ADNI Baseline 
images.  Since our segmentation protocol differs from what can be segmented at 
ADNI resolution a direct comparison is not appropriate.  Instead, we validate by 
showing how well our segmentation volumes correlate with existing automated/manual 
segmentation volumes. \todo{Also show disease prediction?}
\\ \\
{\bf Results: }

\item[First Episode SZ Patients Segmentation] \hfill \\ 
{\bf Purpose:} validate MAGeT's general performance.
\\ \\
{\bf Method:} Measure performance on a different disease dataset. We use FEP SZ patients and controls, 
using two different atlas sets: a manual hippocampal segmentation of patients, and Winterburn 
atlas set.  We validate the native-atlas set segmentations using Dice's Kappa, 
and the Winterburn-atlas-set segmentations by correlating volumes. 
\\ \\
{\bf Results:} High volume correlation between Winterburn segmentation volumes and ground truth.  (High-ish?) Kappa when using manual segmentations as Atlases.

\end{description}
\end{kframe}

@@

Age~at~baseline~\hfill\tiny{Years}&{\scriptsize 72.2~}{75.5 }{\scriptsize 78.5} &{\scriptsize 71.0~}{77.1 }{\scriptsize 81.4} &{\scriptsize 71.7~}{77.8 }{\scriptsize 81.8} &{\scriptsize 71.5~}{76.6 }{\scriptsize 81.3} &$ F_{2,66}=0.15 ,~ P=0.862 ^{1} $\NN

@@

CDR-SB&{\scriptsize 0.00~}{0.00 }{\scriptsize 0.00} &{\scriptsize 0.75~}{1.50 }{\scriptsize 1.50} &{\scriptsize 4.00~}{4.50 }{\scriptsize 5.00} &{\scriptsize 0.00~}{1.50 }{\scriptsize 4.00} &$ F_{2,66}=294 ,~ P<0.001 ^{1} $\NN
ADAS~13&{\scriptsize  4.67~}{ 5.67 }{\scriptsize 12.34} &{\scriptsize 14.34~}{16.00 }{\scriptsize 20.50} &{\scriptsize 23.83~}{29.00 }{\scriptsize 31.66} &{\scriptsize 10.00~}{16.00 }{\scriptsize 25.33} &$ F_{2,66}=110 ,~ P<0.001 ^{1} $\NN
MMSE&{\scriptsize 28.5~}{29.0 }{\scriptsize 30.0} &{\scriptsize 25.0~}{27.0 }{\scriptsize 28.0} &{\scriptsize 21.0~}{23.0 }{\scriptsize 24.0} &{\scriptsize 24.0~}{27.0 }{\scriptsize 29.0} &$ F_{2,66}=94.2 ,~ P<0.001 ^{1} $\LL

@@

\begin{figure}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}

@@

  \caption{{\bf First Episode Schizophrenic Patients.} Comparison of total HC volumes for MAGeT against manually rated volumes of}
  \label{SZ_volumes}
\end{figure}

Some other text here... 
\end{document}
